---
title: "Week 11: Panel Data Causal Inference"
subtitle: "PS 813 - Causal Inference"
author:
  - name: "Anton Strezhnev"
    affiliation: "University of Wisconsin-Madison"
    email: strezhnev@wisc.edu
date: "March 23, 2026"
date-format: "MMMM D, YYYY"
editor:
  render-on-save: true
format:
  revealjs:
    theme: [default, custom.scss]
    slide-number: true
    overview: true
    navigation-mode: vertical
    progress: false
    controls: true
    link-external-newwindow: true
    footer: "PS 813 - University of Wisconsin - Madison"
    html-math-method:
      method: mathjax
      url: https://cdn.jsdelivr.net/npm/mathjax@3/es5/tex-mml-chtml.js
execute:
  cache: true
code:
  execute:
    - r
    - python
embed-resources: true

---

$$
  \require{cancel}
$$

```{r}
#| echo: false
#| warning: false
#| message: false
library(tidyverse)
library(haven)
library(estimatr)
library(panelView)
library(fixest)
options(digits=3)
```

## Last two weeks

::: {.incremental}

- Differences-in-differences designs for dealing with unobserved confounding.
- Identification under **Parallel trends**

  $$E[Y_{it}(0) - Y_{it^{\prime}}(0) \mid D_i = 1] - E[Y_{it}(0) - Y_{it^{\prime}}(0) \mid D_i = 0]$$
  
- Estimation via **first-differences** regression (Callaway and Sant'Anna, 2021)
  - Can adjust for **time-varying** and **time-invariant** threats to parallel trends.
- Pre-treatment **placebo test** diagnostics
- Be careful with software defaults!

:::

## This week

::: {.incremental}

- In DiD, our identifying assumptions **rule out** outcome-treatment feedback
  - Treatment is **exogenous** conditional on observed and unobserved (but structurally constrained) factors
  - But past outcomes do not directly affect present treatment.
- Not uncommon to have **dynamic** treatment settings where this assumption is violated.
  - Doctors change treatment regimen in response to patient outcomes.
  - Political campaigns adjust their strategy in response to observed poll numbers.
  - Economic growth affects democratization (which affects economic growth)
- In all of these settings, we have $Y_{t-1}$ affecting treatment $D_{t}$
  - And typically also lagged **treatment** $D_{t-1}$ affecting current treatment $D_{t}$
  
:::

## Review: DiD and Strict Exogeneity

::: {.incremental}

- For this section, we're going to work with binary treatment $D_{it}$ observed at each time period $t$
  - For simplicity, we'll consider **two** time periods generically as $t$ and $t-1$
  - Let $D_i = \{D_{i1}, D_{i2}, D_{i3}, \dotsc, D_{iT}\}$ denote the entire treatment **history** for a unit
  
- In DiD, our **parallel trends** assumption is a special case of a particular kind of exogeneity/ignorability assumption
  - We assume that treatment history is **as-good-as-randomly** assigned given observed covariates **and** some latent factor
  - Let $\mathbf{X}_i = \{X_{i1}, X_{i2}, X_{i3}, \dotsc, X_{iT}\}$ denote the unit's observed covariate history
  - Let $\mathbf{U}_i = \{U_{i1}, U_{i2}, U_{i3}, \dotsc, U_{iT}\}$ denote the unit's latent factor history
  
- **Strict exogeneity**

  $$\{Y_{it}(0), Y_{it}(1) \} \perp\!\!\!\perp D_{i} | \mathbf{X}_{i}, \mathbf{U}_i$$

:::

## Strict exogeneity

```{tikz}
#| echo: false
#| fig-align: center
#| fig-dpi: 300
#| out-width: "80%"
#| fig-format: "svg"
#| fig-cap: "Adapted from Xu (2023) 'Causal Inference with Time-Series Cross-Sectional Data: A Reflection'"
\usetikzlibrary{arrows}
\begin{tikzpicture}[scale=2, every node/.style={font=\large}]

% Unobserved row (top) -- like X but gray/dashed
\node[gray] (Ut1) at (0.6, 2) {$U_{t-1}$};
\node[gray] (Ut) at (2.6, 2) {$U_{t}$};

% Treatment row
\node (Dt1) at (0.6, 1.2) {$D_{t-1}$};
\node (Dt) at (2.6, 1.2) {$D_{t}$};

% Outcome row (offset right, more space from D row)
\node (Yt1) at (1.3, 0.4) {$Y_{t-1}$};
\node (Yt) at (3.3, 0.4) {$Y_{t}$};

% Observed covariate row (bottom)
\node (Xt1) at (0.6, -0.4) {$X_{t-1}$};
\node (Xt) at (2.6, -0.4) {$X_{t}$};

% Continuation nodes (each row gets its own, no Y continuation)
\node (pastU) at (-0.6, 2) {\textellipsis};
\node (pastD) at (-0.6, 1.2) {\textellipsis};
\node (pastX) at (-0.6, -0.4) {\textellipsis};
\node (futU) at (3.8, 2) {\textellipsis};
\node (futD) at (3.8, 1.2) {\textellipsis};
\node (futX) at (3.8, -0.4) {\textellipsis};

% U -> D within period (gray dashed -- unobserved)
\draw[->, >=stealth, thick, gray, dashed] (Ut1) -- (Dt1);
\draw[->, >=stealth, thick, gray, dashed] (Ut) -- (Dt);

% U -> Y within period (gray dashed, curved to avoid overlap)
\draw[->, >=stealth, thick, gray, dashed] (Ut1) to[bend left=15] (Yt1);
\draw[->, >=stealth, thick, gray, dashed] (Ut) to[bend left=15] (Yt);

% U persistence across periods (gray dashed)
\draw[->, >=stealth, thick, gray, dashed] (Ut1) -- (Ut);

% X -> D within period
\draw[->, >=stealth, thick] (Xt1) -- (Dt1);
\draw[->, >=stealth, thick] (Xt) -- (Dt);

% X -> Y within period
\draw[->, >=stealth, thick] (Xt1) -- (Yt1);
\draw[->, >=stealth, thick] (Xt) -- (Yt);

% X persistence across periods
\draw[->, >=stealth, thick] (Xt1) -- (Xt);

% D -> Y within period (causal effect of interest)
\draw[->, >=stealth, thick] (Dt1) -- (Yt1);
\draw[->, >=stealth, thick] (Dt) -- (Yt);

% D persistence across periods
\draw[->, >=stealth, thick] (Dt1) -- (Dt);

% Continuation arrows (parallel, each row to its own node)
\draw[->, >=stealth, thick, dotted] (pastU) -- (Ut1);
\draw[->, >=stealth, thick, dotted] (pastD) -- (Dt1);
\draw[->, >=stealth, thick, dotted] (pastX) -- (Xt1);
\draw[->, >=stealth, thick, dotted] (Ut) -- (futU);
\draw[->, >=stealth, thick, dotted] (Dt) -- (futD);
\draw[->, >=stealth, thick, dotted] (Xt) -- (futX);

\end{tikzpicture}
```


## DiD and parallel trends

::: {.incremental}

- If we want to identify the effect of $D_{t}$ on $Y_{t}$ (the "instantaneous" effect), via a standard selection-on-observables design, we would need to adjust for both $X$ and $U$
- But here, we've supposed that $U$ is **unobserved** - can't control for it directly.
  - We need **more assumptions**!
- DiD works by making an additional **functional form** assumption about the latent factor.
  - We assume that we can decompose the latent factor into two additive components: $U_{it} = \alpha_i + \delta_t$
- This gives us the classic **two-way fixed effect** formulation:

  $$Y_{it}(0) = \alpha_i + \delta_t + f(\mathbf{X_i}) + \varepsilon_{it}$$

:::

## DiD and parallel trends

::: {.incremental}

- It's straightforward to see under this functional form assumption and the **strict exogeneity** assumption, we get **parallel trends**

  $$\mathbb{E}[Y_{it}(0) - Y_{it-1}(0) \mid D_{it} = 0, \mathbf{X}] = \mathbb{E}[Y_{it}(0) - Y_{it-1}(0) \mid D_{it} = 1, \mathbf{X}] = \delta_{t} - \delta_{t-1}$$ 

  
- Under strict exogeneity $\text{Cov}(\varepsilon_{it}, D_{it}) = 0$

:::

## Sequential ignorability

::: {.incremental}

- But in many panel data settings, ruling out the arrow from $Y_{t-1}$ to $D_{t}$ is implausible!
  - We'll need a **different** assumption
- **Sequential ignorability** (Robins, 1986; Blackwell and Glynn, 2018)
  - At each time period, treatment is as-if random conditional on observed covariate and outcome history

    $$\{Y_{it}(1), Y_{it}(0)\} \perp\!\!\!\perp  D_{it} \mid X_{it}, Y_{it-1}, D_{it-1}$$
  - We can relax this by conditioning on a **longer** history of outcomes, treatments and covariates.

:::

## Sequential ignorability

```{tikz}
#| echo: false
#| fig-align: center
#| fig-dpi: 300
#| out-width: "80%"
#| fig-format: "svg"
#| fig-cap: "Adapted from Blackwell and Glynn (2018) 'How to Make Causal Inferences with Time-Series Cross-Sectional Data under Selection on Observables'"
\usetikzlibrary{arrows}
\begin{tikzpicture}[scale=2, every node/.style={font=\large}]

% Treatment row
\node (Dt1) at (0.6, 1.2) {$D_{t-1}$};
\node (Dt) at (2.6, 1.2) {$D_{t}$};

% Outcome row (offset right)
\node (Yt1) at (1.3, 0.4) {$Y_{t-1}$};
\node (Yt) at (3.3, 0.4) {$Y_{t}$};

% Observed covariate row (bottom)
\node (Xt1) at (0.6, -0.4) {$X_{t-1}$};
\node (Xt) at (2.6, -0.4) {$X_{t}$};

% Continuation nodes (all aligned vertically)
\node (pastD) at (-0.9, 1.2) {\textellipsis};
\node (pastY) at (-0.9, 0.4) {\textellipsis};
\node (pastX) at (-0.9, -0.4) {\textellipsis};
\node (futD) at (4.5, 1.2) {\textellipsis};
\node (futY) at (4.5, 0.4) {\textellipsis};
\node (futX) at (4.5, -0.4) {\textellipsis};

% X -> D within period
\draw[->, >=stealth, thick] (Xt1) -- (Dt1);
\draw[->, >=stealth, thick] (Xt) -- (Dt);

% X -> Y within period
\draw[->, >=stealth, thick] (Xt1) -- (Yt1);
\draw[->, >=stealth, thick] (Xt) -- (Yt);

% D -> Y within period (causal effect of interest)
\draw[->, >=stealth, thick] (Dt1) -- (Yt1);
\draw[->, >=stealth, thick] (Dt) -- (Yt);

% D persistence across periods
\draw[->, >=stealth, thick] (Dt1) -- (Dt);

% X persistence across periods
\draw[->, >=stealth, thick] (Xt1) -- (Xt);

% Cross-period feedback arrows (new relative to strict exogeneity)
\draw[->, >=stealth, thick] (Yt1) -- (Yt);
\draw[->, >=stealth, thick] (Yt1) to[bend right=15] (Dt);
\draw[->, >=stealth, thick] (Dt1) to[bend left=15] (Yt);
\draw[->, >=stealth, thick] (Dt1) -- (Xt);
\draw[->, >=stealth, thick] (Xt1) to[bend right=15] (Dt);

% Continuation arrows (parallel, each row to its own node)
\draw[->, >=stealth, thick, dotted] (pastD) -- (Dt1);
\draw[->, >=stealth, thick, dotted] (pastY) -- (Yt1);
\draw[->, >=stealth, thick, dotted] (pastX) -- (Xt1);
\draw[->, >=stealth, thick, dotted] (Dt) -- (futD);
\draw[->, >=stealth, thick, dotted] (Yt) -- (futY);
\draw[->, >=stealth, thick, dotted] (Xt) -- (futX);

\end{tikzpicture}
```

## Causal estimands

::: {.incremental}

- We now have to consider potential outcomes in terms of the entire **treatment history** assigned up to time $t$

  $$Y_{it}(d_1, d_2, \dotsc, d_t) = Y_{it}(d_{1:t})$$
  
- Partition this into the **contemporaneous** treatment $d_t$ and past values $d_1, d_2, \dotsc, d_{t-1}$

  $$Y_{it}(d_{1:t}) = Y_{it}(d_{t}; d_{1:t-1})$$
  
- The **Contemporaneous Effect** of Treatment

  $$Y_{it}(1; d_{1:t-1}) - Y_{it}(0; d_{1:t-1}) = Y_{it}(1) - Y_{it}(0)$$
  
- Later on when we talk about **mediation**, we'll introduce other kinds of estimands for **cumulative** effects over time.

:::

## Estimation of treatment effects

::: {.incremental}

- Under sequential ignorability, we can motivate estimation of the **contemporaneous** effect of treatment using an Autoregressive Distributed Lag (ADL) model framework.
  - Adjust for some number of **lagged** outcomes (autoregressive)
  - Adjust for some number of **lagged** treatments (distributed lag)
  
- For example, a one-lag linear model

  $$Y_{it} = X_{it}^\prime\beta + \alpha Y_{it-1} + \tau_0 D_{it} + \tau_1 D_{it-1} + \epsilon_{it}$$

- If we believe the parametric assumptions behind this model (e.g. constant effects), $\tau_0$ identifies the **average** contemporaneous effect of treatment at time $t$.
- But does $\tau_1$ identify the effect **one period out**?
  - No! We have **post-treatment** bias **(Glynn and Blackwell, 2018)**.
  - This is the **Table 2 fallacy** all over again.

:::

## Sequential ignorability

```{tikz}
#| echo: false
#| fig-align: center
#| fig-dpi: 300
#| out-width: "80%"
#| fig-format: "svg"
#| fig-cap: "Adapted from Blackwell and Glynn (2018) 'How to Make Causal Inferences with Time-Series Cross-Sectional Data under Selection on Observables'"
\usetikzlibrary{arrows}
\begin{tikzpicture}[scale=2, every node/.style={font=\large}]

% Treatment row
\node (Dt1) at (0.6, 1.2) {$D_{t-1}$};
\node (Dt) at (2.6, 1.2) {$D_{t}$};

% Outcome row (offset right)
\node (Yt1) at (1.3, 0.4) {$Y_{t-1}$};
\node (Yt) at (3.3, 0.4) {$Y_{t}$};

% Observed covariate row (bottom)
\node (Xt1) at (0.6, -0.4) {$X_{t-1}$};
\node (Xt) at (2.6, -0.4) {$X_{t}$};

% Continuation nodes (each row gets its own)
\node (pastD) at (-0.6, 1.2) {\textellipsis};
\node (pastX) at (-0.6, -0.4) {\textellipsis};
\node (futD) at (3.8, 1.2) {\textellipsis};
\node (futX) at (3.8, -0.4) {\textellipsis};

% X -> D within period
\draw[->, >=stealth, thick] (Xt1) -- (Dt1);
\draw[->, >=stealth, thick] (Xt) -- (Dt);

% X -> Y within period
\draw[->, >=stealth, thick] (Xt1) -- (Yt1);
\draw[->, >=stealth, thick] (Xt) -- (Yt);

% D -> Y within period (causal effect of interest)
\draw[->, >=stealth, thick] (Dt1) -- (Yt1);
\draw[->, >=stealth, thick] (Dt) -- (Yt);

% D persistence across periods
\draw[->, >=stealth, thick] (Dt1) -- (Dt);

% X persistence across periods
\draw[->, >=stealth, thick] (Xt1) -- (Xt);

% Cross-period feedback arrows (new relative to strict exogeneity)
\draw[->, >=stealth, thick] (Yt1) -- (Yt);
\draw[->, >=stealth, thick] (Yt1) to[bend right=15] (Dt);
\draw[->, >=stealth, thick] (Dt1) to[bend left=15] (Yt);
\draw[->, >=stealth, thick] (Dt1) -- (Xt);
\draw[->, >=stealth, thick] (Xt1) to[bend right=15] (Dt);

% Continuation arrows (parallel, each row to its own node)
\draw[->, >=stealth, thick, dotted] (pastD) -- (Dt1);
\draw[->, >=stealth, thick, dotted] (pastX) -- (Xt1);
\draw[->, >=stealth, thick, dotted] (Dt) -- (futD);
\draw[->, >=stealth, thick, dotted] (Xt) -- (futX);

\end{tikzpicture}
```

## LDV vs. DiD

::: {.incremental}

- We'll talk about how to estimate these long-term and cumulative effects under **sequential ignorability** when we get to mediation in a few weeks.
  - The relevant estimands and estimators come from this literature.
  - Future treatment and covariates are **mediators** of past treatment!
- But let's consider the **simple setting** with two time periods and two treatment groups
  - Everyone is under **control** at time $t-1$, some units are treated at time $t$

- **Two** possible ways to estimate the treatment effect.
  1. **Difference-in-differences** - Regress $Y_{it} - Y_{it-1}$ on $D_{it}$
  2. **Lagged-DV adjustment** - Regress $Y_{it}$ on $D_{it}$ and $Y_{it-1}$

:::

## Controlling for LDV under strict exogeneity

::: {.incremental}

- What happens if we control for the lagged dependent variable under strict exogeneity?
  - We have a **reversion-to-the-mean** problem
- What we **actually** want to do is adjust for the **expected** lagged outcome.
  - But the actual observed outcome contains **random error**
  - $Y_{it-1}$ can be high because of the **latent factor** or the **error**
- This is another case of **endogenous selection bias** or **collider bias**
  - If treated units tend to be higher in their outcomes than control (on average), then control units with similar **observed** outcomes are **outliers**
  - We'd expect the subsequent outcome to be closer to the mean!

:::

## Simulation: Controlling for LDV

::: {.incremental}

- To see the bias from controlling for lagged $Y$, consider the following DGP:
  - $N = 2000$ units, two periods ($t{-}1$ and $t$), true treatment effect $\tau = 0$
  - Time-invariant confounder: $\alpha_i \sim N(0, 1)$
  - Treatment correlated with confounder: $D_i \sim \text{Bernoulli}(\text{logit}^{-1}(\alpha_i))$
  - Outcome at $t{-}1$: $Y_{it-1} = \alpha_i + \varepsilon_{it-1}$, where $\varepsilon_{it-1} \sim N(0, \sigma_\varepsilon^2)$
  - Outcome at $t$: $Y_{it} = \alpha_i + \beta D_i + \varepsilon_{it}$, where $\varepsilon_{it} \sim N(0, \sigma_\varepsilon^2)$
- **Parallel trends** holds by construction: $\alpha_i$ is time-invariant
- Vary $\sigma_\varepsilon \in \{0.5, 1, 2, 4, 8\}$ to control how **noisy** the proxy $Y_{it-1}$ is for $\alpha_i$
- Compare three estimators: simple difference-in-means post-treatment, DiD, and LDV

:::

## Simulation: LDV bias from measurement error

```{r}
#| echo: true
#| warning: false
#| message: false
#| code-fold: true
#| fig-align: center
#| fig-width: 10
#| fig-height: 5
#| cache: true

set.seed(53703)
N <- 2000
n_sims <- 500
noise_levels <- c(0.5, 1, 2, 4, 8)

sim_results <- map_dfr(noise_levels, function(sigma_e) {
  map_dfr(1:n_sims, function(s) {
    # Time-invariant confounder
    alpha_i <- rnorm(N, 0, 1)
    # Treatment correlated with confounder
    D_i <- rbinom(N, 1, plogis(alpha_i))
    # True treatment effect is ZERO
    beta <- 0
    # Pre and post outcomes: parallel trends holds
    Y_pre <- alpha_i + rnorm(N, 0, sigma_e)
    Y_post <- alpha_i + beta * D_i + rnorm(N, 0, sigma_e)

    # Difference-in-means (biased by alpha_i)
    dim_est <- coef(lm(Y_post ~ D_i))["D_i"]
    # DiD (unbiased)
    did_est <- coef(lm(I(Y_post - Y_pre) ~ D_i))["D_i"]
    # LDV (biased -- errors in variables)
    ldv_est <- coef(lm(Y_post ~ D_i + Y_pre))["D_i"]

    tibble(sigma_e = sigma_e, sim = s,
           `Diff-in-means` = dim_est, DiD = did_est, LDV = ldv_est)
  })
})

sim_summary <- sim_results %>%
  pivot_longer(cols = c(`Diff-in-means`, DiD, LDV),
               names_to = "Estimator", values_to = "estimate") %>%
  group_by(sigma_e, Estimator) %>%
  summarize(mean_est = mean(estimate),
            ci_low = quantile(estimate, 0.025),
            ci_high = quantile(estimate, 0.975), .groups = "drop")

sim_summary$Estimator <- factor(sim_summary$Estimator,
                                levels = c("Diff-in-means", "LDV", "DiD"))

ggplot(sim_summary, aes(x = factor(sigma_e), y = mean_est,
                        color = Estimator, group = Estimator)) +
  geom_point(size = 3, position = position_dodge(width = 0.4)) +
  geom_errorbar(aes(ymin = ci_low, ymax = ci_high),
                width = 0.3, position = position_dodge(width = 0.4)) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(x = expression(paste("Noise in pre-treatment outcome (", sigma[epsilon], ")")),
       y = "Estimate",
       title = "LDV bias under the DiD model (true effect = 0)",
       subtitle = "More noise in Y(pre) = worse proxy for confounder = more residual bias") +
  scale_color_manual(values = c("Diff-in-means" = "gray60",
                                "LDV" = "dodgerblue",
                                "DiD" = "indianred")) +
  theme_bw() + theme(legend.position = "bottom")
```


## Simulation: What if outcomes are autocorrelated?

::: {.incremental}

- In the previous simulation, $\varepsilon_{it-1}$ and $\varepsilon_{it}$ are **independent** — the only thing linking outcomes over time is $\alpha_i$
- In practice, outcomes are often **autocorrelated**: a unit that is high at $t{-}1$ (beyond $\alpha_i$) tends to stay high at $t$
- Modified DGP: let $\varepsilon_{it} = \rho \cdot \varepsilon_{it-1} + \nu_i$ where $\nu_i \sim N(0, \sigma_\varepsilon^2)$
  - When $\rho > 0$, $Y_{it-1}$ is a **better proxy** for $Y_{it}$'s non-treatment component
  - LDV should perform better because controlling for $Y_{it-1}$ captures more of the variation
- We fix $\sigma_\varepsilon = 2$ and vary $\rho \in \{0, 0.25, 0.5, 0.75, 0.95\}$

:::

## Simulation: Autocorrelation reduces LDV bias

```{r}
#| echo: true
#| warning: false
#| message: false
#| code-fold: true
#| fig-align: center
#| fig-width: 10
#| fig-height: 5
#| cache: true

set.seed(53703)
N <- 2000
n_sims <- 500
sigma_e <- 2
rho_levels <- c(0, 0.25, 0.5, 0.75, 0.95)

sim_results_rho <- map_dfr(rho_levels, function(rho) {
  map_dfr(1:n_sims, function(s) {
    alpha_i <- rnorm(N, 0, 1)
    D_i <- rbinom(N, 1, plogis(alpha_i))
    beta <- 0

    eps_pre <- rnorm(N, 0, sigma_e)
    eps_post <- rho * eps_pre + rnorm(N, 0, sigma_e)

    Y_pre <- alpha_i + eps_pre
    Y_post <- alpha_i + beta * D_i + eps_post

    dim_est <- coef(lm(Y_post ~ D_i))["D_i"]
    did_est <- coef(lm(I(Y_post - Y_pre) ~ D_i))["D_i"]
    ldv_est <- coef(lm(Y_post ~ D_i + Y_pre))["D_i"]

    tibble(rho = rho, sim = s,
           `Diff-in-means` = dim_est, DiD = did_est, LDV = ldv_est)
  })
})

sim_summary_rho <- sim_results_rho %>%
  pivot_longer(cols = c(`Diff-in-means`, DiD, LDV),
               names_to = "Estimator", values_to = "estimate") %>%
  group_by(rho, Estimator) %>%
  summarize(mean_est = mean(estimate),
            ci_low = quantile(estimate, 0.025),
            ci_high = quantile(estimate, 0.975), .groups = "drop")

sim_summary_rho$Estimator <- factor(sim_summary_rho$Estimator,
                                    levels = c("Diff-in-means", "LDV", "DiD"))

ggplot(sim_summary_rho, aes(x = factor(rho), y = mean_est,
                            color = Estimator, group = Estimator)) +
  geom_point(size = 3, position = position_dodge(width = 0.4)) +
  geom_errorbar(aes(ymin = ci_low, ymax = ci_high),
                width = 0.3, position = position_dodge(width = 0.4)) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(x = expression(paste("Autocorrelation (", rho, ")")),
       y = "Estimate",
       title = expression(paste("LDV bias with autocorrelated errors (", sigma[epsilon], " = 2, true effect = 0)")),
       subtitle = "More autocorrelation = lagged Y is better proxy = less LDV bias") +
  scale_color_manual(values = c("Diff-in-means" = "gray60",
                                "LDV" = "dodgerblue",
                                "DiD" = "indianred")) +
  theme_bw() + theme(legend.position = "bottom")
```


## Remember the DAG!

```{tikz}
#| echo: false
#| fig-align: center
#| fig-dpi: 300
#| out-width: "80%"
#| fig-format: "svg"
#| fig-cap: "Adapted from Xu (2023) 'Causal Inference with Time-Series Cross-Sectional Data: A Reflection'"
\usetikzlibrary{arrows}
\begin{tikzpicture}[scale=2, every node/.style={font=\large}]

% Unobserved row (top) -- like X but gray/dashed
\node[gray] (Ut1) at (0.6, 2) {$U_{t-1}$};
\node[gray] (Ut) at (2.6, 2) {$U_{t}$};

% Treatment row
\node (Dt1) at (0.6, 1.2) {$D_{t-1}$};
\node (Dt) at (2.6, 1.2) {$D_{t}$};

% Outcome row (offset right, more space from D row)
\node (Yt1) at (1.3, 0.4) {$Y_{t-1}$};
\node (Yt) at (3.3, 0.4) {$Y_{t}$};

% Observed covariate row (bottom)
\node (Xt1) at (0.6, -0.4) {$X_{t-1}$};
\node (Xt) at (2.6, -0.4) {$X_{t}$};

% Continuation nodes (each row gets its own, no Y continuation)
\node (pastU) at (-0.6, 2) {\textellipsis};
\node (pastD) at (-0.6, 1.2) {\textellipsis};
\node (pastX) at (-0.6, -0.4) {\textellipsis};
\node (futU) at (3.8, 2) {\textellipsis};
\node (futD) at (3.8, 1.2) {\textellipsis};
\node (futX) at (3.8, -0.4) {\textellipsis};

% U -> D within period (gray dashed -- unobserved)
\draw[->, >=stealth, thick, gray, dashed] (Ut1) -- (Dt1);
\draw[->, >=stealth, thick, gray, dashed] (Ut) -- (Dt);

% U -> Y within period (gray dashed, curved to avoid overlap)
\draw[->, >=stealth, thick, gray, dashed] (Ut1) to[bend left=15] (Yt1);
\draw[->, >=stealth, thick, gray, dashed] (Ut) to[bend left=15] (Yt);

% U persistence across periods (gray dashed)
\draw[->, >=stealth, thick, gray, dashed] (Ut1) -- (Ut);

% X -> D within period
\draw[->, >=stealth, thick] (Xt1) -- (Dt1);
\draw[->, >=stealth, thick] (Xt) -- (Dt);

% X -> Y within period
\draw[->, >=stealth, thick] (Xt1) -- (Yt1);
\draw[->, >=stealth, thick] (Xt) -- (Yt);

% X persistence across periods
\draw[->, >=stealth, thick] (Xt1) -- (Xt);

% D -> Y within period (causal effect of interest)
\draw[->, >=stealth, thick] (Dt1) -- (Yt1);
\draw[->, >=stealth, thick] (Dt) -- (Yt);

% D persistence across periods
\draw[->, >=stealth, thick] (Dt1) -- (Dt);

% Continuation arrows (parallel, each row to its own node)
\draw[->, >=stealth, thick, dotted] (pastU) -- (Ut1);
\draw[->, >=stealth, thick, dotted] (pastD) -- (Dt1);
\draw[->, >=stealth, thick, dotted] (pastX) -- (Xt1);
\draw[->, >=stealth, thick, dotted] (Ut) -- (futU);
\draw[->, >=stealth, thick, dotted] (Dt) -- (futD);
\draw[->, >=stealth, thick, dotted] (Xt) -- (futX);

\end{tikzpicture}
```



## Illustration: The Effects of the Shale Shock

- Let's return to our example of the effect of the "shale shock" on coal county voting for the Republican party (Gazmararian, 2025).
  - Last week we showed concerning evidence of a **parallel trends** violation - coal-heavy regions were already trending Republican prior to 2008.
  - Extrapolating the trend (triple-differences in time) eliminates the observed effect.
- What if, instead of DiD, we **controlled** for lagged Republican party vote share in 2004 and 2000.

## Lagged DV approach: Setup

```{r}
#| echo: true
#| warning: false
#| message: false
#| code-fold: true

dff <- readRDS("assets/data/matched_df.rds")
dff <- dff %>% filter(!is.na(oilgas_prop))

# Create wide-format lagged outcomes
lag_vars <- dff %>%
  select(county_state, year, RepVotesMajorPercent) %>%
  pivot_wider(names_from = year, values_from = RepVotesMajorPercent,
              names_prefix = "Y_")

dff <- dff %>% left_join(lag_vars, by = "county_state")
```

## Lagged DV: Estimates by year

```{r}
#| echo: true
#| warning: false
#| message: false
#| code-fold: true
#| fig-align: center
#| fig-width: 8
#| fig-height: 4.5

# LDV regressions for each post-treatment year
ldv_years <- c(2008, 2012, 2016, 2020)
ldv_results <- map_dfr(ldv_years, function(yr) {
  df_yr <- dff %>% filter(year == yr)
  fit <- estimatr::lm_robust(
    RepVotesMajorPercent ~ treat + Y_2004 + Y_2000 + oilgas_prop,
    data = df_yr, clusters = county_state, se_type = "stata")
  tidy_fit <- broom::tidy(fit) %>% filter(term == "treat")
  tidy_fit$year <- yr
  tidy_fit
})

ggplot(ldv_results, aes(x = year, y = estimate)) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = conf.low, ymax = conf.high), width = 0.5) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_vline(xintercept = 2006, linetype = "dashed", color = "red") +
  labs(x = "Year", y = "Estimate (pp)",
       title = "LDV: Effect of coal county on Republican vote share",
       subtitle = "Controlling for Y[2004], Y[2000], and oil/gas employment") +
  scale_x_continuous(breaks = ldv_years) +
  theme_bw()
```


## DiD: Estimates by year

```{r}
#| echo: true
#| warning: false
#| message: false
#| code-fold: true
#| fig-align: center
#| fig-width: 8
#| fig-height: 4.5

# DiD regressions for each post-treatment year
did_results <- map_dfr(ldv_years, function(yr) {
  df_yr <- dff %>%
    filter(year == yr) %>%
    mutate(dY = RepVotesMajorPercent - Y_2004)
  fit <- estimatr::lm_robust(
    dY ~ treat + oilgas_prop,
    data = df_yr, clusters = county_state, se_type = "stata")
  tidy_fit <- broom::tidy(fit) %>% filter(term == "treat")
  tidy_fit$year <- yr
  tidy_fit
})

ggplot(did_results, aes(x = year, y = estimate)) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = conf.low, ymax = conf.high), width = 0.5) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_vline(xintercept = 2006, linetype = "dashed", color = "red") +
  labs(x = "Year", y = "Estimate (pp)",
       title = "DiD: Effect of coal county on change in Republican vote share",
       subtitle = "Outcome: Y[t] - Y[2004], controlling for oil/gas employment") +
  scale_x_continuous(breaks = ldv_years) +
  theme_bw()
```

## Comparing LDV and DiD

```{r}
#| echo: false
#| warning: false
#| message: false
#| fig-align: center
#| fig-width: 10
#| fig-height: 4.5

combined <- bind_rows(
  ldv_results %>% mutate(model = "LDV"),
  did_results %>% mutate(model = "DiD")
)

ggplot(combined, aes(x = year, y = estimate, color = model, shape = model)) +
  geom_point(size = 3, position = position_dodge(width = 0.8)) +
  geom_errorbar(aes(ymin = conf.low, ymax = conf.high),
                width = 0.5, position = position_dodge(width = 0.8)) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_vline(xintercept = 2006, linetype = "dashed", color = "red") +
  labs(x = "Year", y = "Estimate (pp)",
       title = "Comparing LDV vs. DiD estimates",
       color = "Model", shape = "Model") +
  scale_x_continuous(breaks = ldv_years) +
  scale_color_manual(values = c("LDV" = "dodgerblue", "DiD" = "indianred")) +
  theme_bw() + theme(legend.position = "bottom")
```


## Placebo test: Pre-treatment hold-out

::: {.incremental}

- Let's construct a **placebo test** for the lagged DV approach.
  - Because we have many pre-treatment periods, we can apply the LDV approach to **pre-treatment** outcomes
- **Placebo 1**: Outcome year 2004, controlling for $Y_{2000}$ and $Y_{1996}$
- **Placebo 2**: Outcome year 2000, controlling for $Y_{1996}$ and $Y_{1992}$
- **Placebo 3**: Outcome year 1996, controlling for $Y_{1992}$ and $Y_{1988}$

:::

## Placebo: LDV estimates

```{r}
#| echo: true
#| warning: false
#| message: false
#| code-fold: true
#| fig-align: center
#| fig-width: 8
#| fig-height: 4.5

# Placebo LDV regressions (2 lagged controls, matching main spec)
placebo_ldv_results <- bind_rows(
  # 2004 outcome, controlling for 2000, 1996
  {
    df_yr <- dff %>% filter(year == 2004)
    fit <- estimatr::lm_robust(
      RepVotesMajorPercent ~ treat + Y_2000 + Y_1996 + oilgas_prop,
      data = df_yr, clusters = county_state, se_type = "stata")
    broom::tidy(fit) %>% filter(term == "treat") %>% mutate(year = 2004)
  },
  # 2000 outcome, controlling for 1996, 1992
  {
    df_yr <- dff %>% filter(year == 2000)
    fit <- estimatr::lm_robust(
      RepVotesMajorPercent ~ treat + Y_1996 + Y_1992 + oilgas_prop,
      data = df_yr, clusters = county_state, se_type = "stata")
    broom::tidy(fit) %>% filter(term == "treat") %>% mutate(year = 2000)
  },
  # 1996 outcome, controlling for 1992, 1988
  {
    df_yr <- dff %>% filter(year == 1996)
    fit <- estimatr::lm_robust(
      RepVotesMajorPercent ~ treat + Y_1992 + Y_1988 + oilgas_prop,
      data = df_yr, clusters = county_state, se_type = "stata")
    broom::tidy(fit) %>% filter(term == "treat") %>% mutate(year = 1996)
  }
)

ggplot(placebo_ldv_results, aes(x = year, y = estimate)) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = conf.low, ymax = conf.high), width = 0.5) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(x = "Year", y = "Estimate (pp)",
       title = "Placebo test: LDV estimates in pre-treatment periods",
       subtitle = "Each placebo controls for two prior election outcomes + oil/gas share") +
  scale_x_continuous(breaks = c(1996, 2000, 2004)) +
  theme_bw()
```

## Bracketing: Angrist and Pischke (2009)

::: {.incremental}

- **Angrist and Pischke (2009, Ch. 5.4)** show that LDV and DiD estimates **bracket** the true causal effect if **one** of the two identifying assumptions holds.
  - Focus on the simple linear model case - **Ding and Li (2019)** extend to the non-parametric setting.

- Two-period DiD. All units under control at time $0$
  - Assume $\tau > 0$ (positive effect)
  
- For exposition, assume time-effects are already removed via differencing and no covariates.

:::

## Bracketing: Case 1 -- DiD is correct, LDV is misspecified

::: {.incremental}

- Suppose the **FE model** is the truth:

  $$Y_{it} = \alpha_i + \tau D_{it} + \varepsilon_{it}$$

  where $\varepsilon_{it}$ is serially uncorrelated and independent of $\alpha_i$ and $D_{it}$. Since $D_{it-1} = 0$: $Y_{it-1} = \alpha_i + \varepsilon_{it-1}$

- You mistakenly estimate the **LDV regression** (controlling for $Y_{it-1}$ instead of $\alpha_i$):

  $$Y_{it} = a + b \cdot Y_{it-1} + \tau^{\text{LDV}} D_{it} + u_{it}$$

:::

## Bracketing: Case 1 -- FWL decomposition

::: {.incremental}

- By FWL, the LDV estimand is $\hat{\tau}^{\text{LDV}} \xrightarrow{p} \frac{\text{Cov}(Y_{it}, \tilde{D}_{it})}{\text{Var}(\tilde{D}_{it})}$
  - $\tilde{D}_{it} = D_{it} - \hat{\gamma} Y_{it-1}$ is the residual from the auxiliary regression of $D_{it}$ on $Y_{it-1}$

- To evaluate this, substitute $\alpha_i = Y_{it-1} - \varepsilon_{it-1}$ into the true model:

  $$Y_{it} = \alpha_i + \tau D_{it} + \varepsilon_{it} = (Y_{it-1} - \varepsilon_{it-1}) + \tau D_{it} + \varepsilon_{it}$$

  $$\Rightarrow Y_{it} = Y_{it-1} + \tau D_{it} + (\varepsilon_{it} - \varepsilon_{it-1})$$

:::

## Bracketing: Case 1 -- Deriving the bias

::: {.incremental}

- Substitute in and use properties of covariances

  $$\frac{\text{Cov}(Y_{it}, \tilde{D}_{it})}{\text{Var}(\tilde{D}_{it})} = \frac{\text{Cov}(Y_{it-1} + \tau D_{it} + \varepsilon_{it} - \varepsilon_{it-1},\; \tilde{D}_{it})}{\text{Var}(\tilde{D}_{it})}$$
  
- $\text{Cov}(Y_{it-1}, \tilde{D}_{it}) = 0$ by construction (FWL residualizes out $Y_{it-1}$)
- $\text{Cov}(D_{it}, \tilde{D}_{it}) = \text{Var}(\tilde{D}_{it})$ 
- $\text{Cov}(\varepsilon_{it}, \tilde{D}_{it}) = 0$ since $\varepsilon_{it} \perp\!\!\!\perp (D_{it}, Y_{it-1})$
- But $\text{Cov}(\varepsilon_{it-1}, \tilde{D}_{it}) = \text{Cov}(\varepsilon_{it-1}, D_{it}) - \hat{\gamma}\text{Cov}(\varepsilon_{it-1}, Y_{it-1})$
  - And since $Y_{it-1} = \alpha_i + \varepsilon_{it-1}$...
  - $\text{Cov}(\varepsilon_{it-1}, \tilde{D}_{it})  = -\hat{\gamma}\sigma^2_{\varepsilon}$

- So: $\hat{\tau}^{\text{LDV}} \xrightarrow{p} \tau + \frac{\hat{\gamma} \sigma^2_\varepsilon}{\text{Var}(\tilde{D}_{it})}$, with bias having the **same sign** as $\hat{\gamma}$ (the sign of selection)

:::

## Bracketing: Case 1 -- Why is LDV biased?

::: {.incremental}

- $Y_{it-1} = \alpha_i + \varepsilon_{it-1}$ is a **noisy proxy** for the confounder $\alpha_i$
  - Controlling for a noisy proxy does not fully remove confounding
- With **positive selection** (high outcome $\leadsto$ more likely to be treated) ($\hat{\gamma} > 0$)
  - LDV **overestimates** $\tau$ 
- With **negative selection** ($\hat{\gamma} < 0$)
  - LDV **underestimates** $\tau$

:::

## Bracketing: Case 2 -- LDV is correct, DiD is misspecified

::: {.incremental}

- Now suppose the **LDV model** is the truth:

  $$Y_{it} = \alpha + \theta Y_{it-1} + \tau D_{it} + \varepsilon_{it}$$

- Where $\varepsilon_{it}$ is serially uncorrelated, $0 < \theta < 1$ (stationary)

- You mistakenly estimate the **first-differenced regression** (no control for $Y_{it-1}$):

  $$\Delta Y_{it} = a + \tau^{\text{DiD}} D_{it} + u_{it}$$

  where $\Delta Y_{it} = Y_{it} - Y_{it-1}$ and $D_{it-1} = 0$ for everyone

:::

## Bracketing: Case 2 -- FWL decomposition

::: {.incremental}

- The DiD estimand is $\hat{\tau}^{\text{DiD}} \xrightarrow{p} \frac{\text{Cov}(\Delta Y_{it}, D_{it})}{\text{Var}(D_{it})}$

- Subtract $Y_{it-1}$ from both sides of the true model:

  $$Y_{it} - Y_{it-1} = \alpha + (\theta - 1)Y_{it-1} + \tau D_{it} + \varepsilon_{it}$$

- So $\Delta Y_{it} = \alpha + (\theta - 1)Y_{it-1} + \tau D_{it} + \varepsilon_{it}$. Substituting:

  $$\frac{\text{Cov}(\Delta Y_{it}, D_{it})}{\text{Var}(D_{it})} = \frac{\text{Cov}(\alpha + (\theta - 1)Y_{it-1} + \tau D_{it} + \varepsilon_{it}, D_{it})}{\text{Var}(D_{it})} $$

  $$\frac{\text{Cov}(\Delta Y_{it}, D_{it})}{\text{Var}(D_{it})} = \tau + (\theta - 1)\frac{\text{Cov}(Y_{it-1}, D_{it})}{\text{Var}(D_{it})} + \frac{\text{Cov}(\varepsilon_{it}, D_{it})}{\text{Var}(D_{it})}$$

- $\text{Cov}(\varepsilon_{it}, D_{it}) = 0$ by assumption (LDV model has no unobserved confounding)

:::

## Bracketing: Case 2 -- Deriving the bias

::: {.incremental}

- Therefore:

  $$\hat{\tau}^{\text{DiD}} \xrightarrow{p} \tau + (\theta - 1)\frac{\text{Cov}(Y_{it-1}, D_{it})}{\text{Var}(D_{it})}$$

- The bias is $(\theta - 1)\frac{\text{Cov}(Y_{it-1}, D_{it})}{\text{Var}(D_{it})}$
- The bias has the **opposite sign** from the selection:
  - With positive selection ($\text{Cov}(Y_{it-1}, D_{it}) > 0$) and $\theta < 1$: bias is **negative** → DiD underestimates
  - With negative selection: bias is **positive** → DiD overestimates

- Under this model, **LDV is unbiased**: it correctly controls for $Y_{it-1}$

:::

## Bracketing

::: {.incremental}

- Combining both cases with **positive selection** ($\text{Cov}(\alpha_i, D_i) > 0$):
  - If strict exogeneity is correct: LDV overestimates, DiD is exact
  - If sequential ignorability is correct: DiD underestimates, LDV is exact

- The true effect is **bracketed** between the two estimates if **one** of these two assumptions is true.

- This is a useful result, but important to be aware of its limitations
  - It guarantees nothing if **neither assumption** is true!
  
- In many cases, we're skeptical of **parallel trends** but we don't think **sequential ignorability** is right either
  - Rather, we think the unobserved latent factor structure might be more complicated than $U_{it} = \alpha_i + \delta_t$
  - What do we do then?
  
:::



