---
title: "Week 3: Experiments - Part 2"
subtitle: "PS 813 - Causal Inference"
author:
  - name: "Anton Strezhnev"
    affiliation: "University of Wisconsin-Madison"
    email: strezhnev@wisc.edu
date: "February 2, 2026"
date-format: "MMMM D, YYYY"
editor:
  render-on-save: true
format:
  revealjs:
    theme: [default, custom.scss]
    slide-number: true
    overview: true
    navigation-mode: vertical
    progress: false
    controls: true
    link-external-newwindow: true
    footer: "PS 813 - University of Wisconsin - Madison"
    html-math-method:
      method: mathjax
      url: https://cdn.jsdelivr.net/npm/mathjax@3/es5/tex-mml-chtml.js
execute:
  cache: true
code:
  execute:
    - r
    - python
embed-resources: true

---


```{r}
#| echo: false
#| message: false
#| warning: false
library(tidyverse)
library(estimatr)
options(digits=3)
```

## Last week

  - Why randomized experiments work
    - Guarantee treatment $D_i$ is independent of potential outcomes $\{Y_i(1), Y_i(0)\}$

. . .

  - Inference for treatment effects
    - **Fisher**: Can get exact p-values under the sharp null just knowing the
    distribution of treatment assignments.
    - **Neyman**: A conservative variance estimator for the difference-in-means + large-sample asymptotics

    $$\widehat{Var(\hat{\tau})} = \frac{s^2_t}{N_t} + \frac{s_c^2}{N_c}$$

## This week

- One post-experiment use of covariates
  - Balance checking

. . .

- When should we *not* condition on covariates?
  - When they're **post-treatment**!
  - Attrition/Non-compliance
  - Bounds when we *have* to condition
  
. . .
  
- How do we **generalize** from a single experiment?
  - What are the dimensions of **external validity**?
  - What do we have to believe to **transport** a treatment effect?

## Balance checking {.title-slide}

$$
  \require{cancel}
$$

## Balance tests

  - One reason to use covariates even in completely randomized designs is to check
whether the experiment **actually** did what it was supposed to do.

. . .

- Under any randomization scheme that satisfies **ignorability**:

$$X_i {\perp \! \! \! \perp} D_i$$


$$E[X_i | D_i = 1] = E[X_i | D_i = 0]$$

. . .

- If we correctly randomized treatment, then the expectation (and the distribution)
of covariates should be **the same** in treatment and control.

. . .

- But in any given sample, we'll observe a difference just by chance -- how do we
know if this **is a problem**?

## Against balance tests

- One view in some experimentally-focused fields is that you should **never** waste time
checking balance

. . .

- **Senn (1994)** *Statistics in Medicine*
  1. Randomization ensures balance over all randomizations.
  2. Observing any particular imbalance in our sample doesn't disprove 1

. . .

- What if we did screw up? What if the randomization software had a bug? What if there was
some implementation issue?
  - Maybe a balance test helps here?

. . .

- But how would we interpret a "failed" balance test?
  - A reason to go back and check the randomization process - if we think that it *did* actually work
  as intended, we may just have gotten unlucky.
  - Might want a stricter threshold than $p < .05$ if we **really** believe treatment was randomized.


## Against balance tests

- Another more nuanced argument is that you shouldn't use balance tests to decide
whether to **include or not include** covariates in the analysis.

. . .

- **Mutz, Pemantle and Pham (2019)**
  - Sometimes, researchers run a lot of univariate balance tests in an experiment.
  - If a test for some covariate fails, they include that covariate in adjustment.
  - This process risks raising false-positive rates through researcher "degrees-of-freedom"

. . .

- This is a correct argument, but is less an argument against *balance testing* per-se and
more against ad-hoc or data-dependent covariate choices.

. . .

- When we talk about covariates in experiments next week, we'll emphasize the importance of **ex-ante** choices.
  - Don't decide on **what** to include in your analysis based on tests conducted **after the experiment is run** and using **information from the experiment outcomes**
  
## Example: Broockman and Kalla (2023)

- **Broockman and Kalla (2023)** "Consuming cross-cutting media causes learning and moderates attitudes: A field experiment with Fox News viewers"
  - Sample of **763** individuals in **695** households who regularly watch FOX News
  - **Treatment**: Incentivized to watch CNN instead of Fox News for one month (**304** individuals)
  - **Control**: No incentive (continue normal viewing habits) (**459** individuals)

. . .

- Randomization was done at the household level, stratified by baseline characteristics
- Let's check balance on pre-treatment covariates!

## Broockman and Kalla: Loading the data

```{r bk-load}
#| echo: true
#| message: false
#| warning: false
library(haven)
library(cobalt)

# Load the Broockman and Kalla data
bk <- read_dta("assets/data/primary_dataset_unstd.dta")

# Treatment variable: treat (1 = CNN incentive, 0 = control)
table(bk$treat)
```

```{r bk-covariates}
#| echo: true
# Key pre-treatment covariates from baseline survey (t1_) and voter file (vf_)
baseline_covs <- c(
  # Demographics from voter file
  "vf_age",                    # Age
  # Baseline survey measures (t1_ = time 1, pre-treatment)
  "t1_pid7",                   # 7-point party ID (1 = Strong Dem to 7 = Strong Rep)
  "t1_ideo_self",              # Self-reported ideology
  "t1_therm_trump",            # Feeling thermometer: Trump
  "t1_therm_biden",            # Feeling thermometer: Biden
  "t1_therm_fox",              # Feeling thermometer: Fox News
  "t1_therm_cnn",              # Feeling thermometer: CNN
  "t1_trust_fox",              # Trust in Fox News
  "t1_trust_cnn",              # Trust in CNN
  # Pre-treatment TV viewership (from set-top box data)
  "pre_treat_fox_minutes",     # Minutes watching Fox pre-treatment
  "pre_treat_cnn_minutes"      # Minutes watching CNN pre-treatment
)
```

## Broockman and Kalla: Aggregate to household level

```{r bk-aggregate}
#| echo: true
# Randomization was at the household level, so aggregate covariates
bk_hh <- bk %>%
  group_by(hh_id) %>%
  summarize(
    treat = first(treat),
    # Take the mean of covariates within household
    across(all_of(baseline_covs), ~mean(.x, na.rm = TRUE))
  ) %>%
  ungroup()

# Check: 695 households
nrow(bk_hh)

# Check treatment assignments
table(bk_hh$treat)
```

## Broockman and Kalla: Balance table

```{r bk-balance}
#| echo: true
#| output-location: slide
# Use cobalt to create a balance table (at household level)
bal_tab <- bal.tab(
  x = bk_hh %>% select(all_of(baseline_covs)),   # Covariate data
  treat = bk_hh$treat,           # Treatment indicator
  binary = "std",                # Standardize binary variables
  continuous = "std",            # Standardize continuous variables
  s.d.denom = "pooled"           # Use pooled SD for standardization
)
print(bal_tab)
```

## Broockman and Kalla: Love plot

```{r bk-love, message = F}
#| echo: false
#| fig-width: 10
#| fig-height: 6
#| fig-align: center
# Create a Love plot to visualize balance (at household level)
love.plot(
  x = bk_hh %>% select(all_of(baseline_covs)),
  treat = bk_hh$treat,
  binary = "std",
  s.d.denom = "pooled",
  thresholds = c(m = 0.1),      # Show threshold at 0.1 SMD
  var.order = "unadjusted",
  abs = TRUE,                    # Show absolute values
  title = "Covariate Balance: Broockman & Kalla (2023)"
)
```

## Multiple testing

- We could put a simple difference-in-means hypothesis test on each of these covariates.
  - But with enough covariates, we'd find a $p < .05$ just by chance!

```{r, echo=T}
lm_robust(t1_therm_trump ~ treat, data=bk_hh)
```

. . . 

- Is there a quick way to deal with multiple testing in balance checks?
  - Could correct for multiple comparisons...
  - ...or construct a **single** test.
  - How? With randomization inference!
  
## Permutation/Randomization Testing for Balance

- **Step 1:** Run a regression predicting treatment using all the covariates. Store the F-statistic (or any statistic summarizing "goodness of fit")

```{r, echo=T}
treat_reg <- lm_robust(treat ~ vf_age + t1_pid7 + t1_ideo_self + t1_therm_trump + t1_therm_biden +
                         t1_therm_fox + t1_therm_cnn + t1_trust_fox + t1_trust_cnn + pre_treat_fox_minutes +
                         pre_treat_cnn_minutes, data=bk_hh)
tstat_obs <- treat_reg$fstatistic[1]
```

## Permutation/Randomization Testing for Balance

- **Step 2:** Permute the treatment assignment based on the known assignment scheme
- **Step 3:** Calculate the test statistic under alternative assignments

```{r, echo=T}
set.seed(53706)
iterations <- 10000
tstat_null <- rep(NA, iterations)
for (i in 1:iterations){
  bk_hh$treat_perm <- sample(bk_hh$treat)
  treat_reg_null <- lm_robust(treat_perm ~ vf_age + t1_pid7 + t1_ideo_self + t1_therm_trump + t1_therm_biden +
                         t1_therm_fox + t1_therm_cnn + t1_trust_fox + t1_trust_cnn + pre_treat_fox_minutes +
                         pre_treat_cnn_minutes, data=bk_hh)
  tstat_null[i] <- treat_reg_null$fstatistic[1]
}
```

## Permutation/Randomization Testing for Balance

- **Step 4:** Compare the observed statistic to the permuted null distribution

```{r, echo=T}
mean(tstat_null > tstat_obs)
```

## Guidelines for balance testing

- Testing for balance to assess whether randomization occurred as intended: [Good]{.green}
  - Careful with multiple testing/false positives
  - $p < .05$ is probably too low of a threshold, but you should probably be concerned if $p < 1 \times 10^{-6}$
  - What do do if a balance check fails? Check your experiment!

. . .

- Testing for balance to pick which covariates to adjust for: [Bad]{.red}
  - "Garden of forking paths"
  - You should choose covariates ex-ante (even if not blocking)
  - Pick covariates that predict $Y$ - balance checks are the wrong criteria.

## Post-treatment bias {.title-slide}

## Post-treatment covariates

- When talking about covariates, we've emphasized that $X_i$ must be **pre-treatment**
- What happens when we condition on some post-treatment variable (call it $M_i$).
- **Intuition**: $M_i$ is post-treatment. It has potential outcomes: $\{M_i(1), M_i(0)\}$
  - But we can't condition on the *latent* potential outcomes, we only condition on the observed $M_i$
  - This induces a form of enodgenous "selection bias"

. . .

- Many cases in political science
  - Experiments with **non-compliance**
  - **Attention checks** in survey experiments.
  - **Administrative data** (police interactions are only recorded if a stop occurs)
  - **Attrition** induced by treatment (e.g. court proceedings that settle)


## Post-treatment bias

- Let $M_i$ denote the post-treatment covariate. Since it's post-treatment, it has potential outcomes $\{M_i(1), M_i(0)\}$ as
though it were any other outcome.

. . .

- By randomization

$$\{M_i(1), M_i(0)\} {\perp \! \! \! \perp} D_i$$

. . .

- What happens if we take the difference-in-means conditional on $M_i = 1$

$$E[Y_i | D_i = 1, M_i = 1] - E[Y_i | D_i = 0, M_i = 1]$$

. . .

- By consistency:

$$E[Y_i(1) | D_i = 1, M_i(1) = 1] - E[Y_i(0) | D_i = 0, M_i(0) = 1]$$


## Post-treatment bias

- Ignorability gets us

$$E[Y_i(1) |  M_i(1) = 1] - E[Y_i(0) |  M_i(0) = 1]$$

. . .

- Is this the ATE?
  - No! $M_i(1) = 1$ and $M_i(0) = 1$ define *two different subsets of the sample*
- Under what assumptions would we get the ATE?

. . .

- Either:
  1. No individual effect of treatment on $M_i$: $M_i(1) = M_i(0) \text{   } \forall i$
  2. $\{M_i(1), M_i(0)\} {\perp \! \! \! \perp} \{Y_i(1), Y_i(0\}$

. . .

- Neither of these assumptions is guaranteed by an experiment since we don't randomize $M_i$
- Therefore, conditioning on a post-treatment quantity **breaks** the experiment -- now it's an observational study.

## Principal strata

- We can think of the combination of $D_i$ and $M_i$ as defining a "sub-group" of units - these are referred to as "principal strata"
  - They have different names in the literature - one common convention comes from the **compliance** literature
  
Stratum              | $M_i(1)$          |  $M_i(0)$ |
:-------------------:|:-----------------:|:---------:|
"Always-takers"      |  $1$              |  $1$      |
"Never-takers"       |  $0$              |  $0$      |
"Compliers"          |  $1$              |  $0$      |
"Defiers"            |  $0$              |  $1$      |

. . .

- Units with $M_i = 1$ could be any three of these strata. Even observing $D_i$ narrows it down to only two - we can't observe the strata directly.
- Strata aren't necessarily independent of potential outcomes $Y_i(d)$!

## Example: Administrative data

- **Knox, Lowe and Mummolo (2020)** consider the problem of estimating the effect of civilian race on police use of force.
  - Typically, past studies would use administrative data from police departments on stops
  - Compare police use of force among Black civilians who are stopped and white civilians who are stopped.
  - **Problem**: Stops are post-treatment!

. . .

- Define $D_i$ as the treatment (race of civilian), $M_i$ is an indicator for whether a stop occurs, $Y_i$ is severe use of force
- The difference-in-means does not identify the treatment effect - instead, we get:

$$E[Y_i(1) |  M_i(1) = 1] - E[Y_i(0) | M_i(0) = 1]$$


## Bounding the Survivor Average Causal Effect

- Can we say anything about the treatment effect?
  - First - we need to define *which* treatment effect we're interested in to begin with!

. . .

- We can define meaningful causal quantities **conditional** on the principal strata
  - Because these are *pre-treatment*
  
. . .

- For example, the **complier** average causal effect

  $$\tau_{\text{CACE}} = E[Y_i(1) - Y_i(0) | M_i(1) = 1, M_i(0) = 0]$$

- We'll talk more about this when we get to **instrumental variables**

## The Survivor Average Causal Effect

- In the attrition setting, we focus on the **Survivor Average Causal Effect (SACE)**
  - The ATE among those who would be observed irrespective of treatment status

$$\tau_{\text{SACE}} = E[Y_i(1) - Y_i(0) | M_i(1) = 1, M_i(0) = 1]$$

. . .

- Recall that the difference-in-means only gets us 

$$E[Y_i(1) |  M_i(1) = 1] - E[Y_i(0) | M_i(0) = 1]$$

. . . 

- $M_i(1) = 1$ is not the same subset of units as $M_i(0) = 1$
  - $M_i(1) = 1$ includes both **survivors** and the "compliers" 
  - $M_i(0) = 1$ includes both **survivors** and the "defiers"

## Monotonicity

- One approach to bounding from **Lee (2009)** relies on an additional **monotonicity** assumption

$$M_i(1) \ge M_i(0) \text{  } \forall i \quad \text{or} \quad M_i(1) \le M_i(0) \text{  } \forall i$$

. . .

- Treatment affects selection in **only one direction**
  - Either treatment can only *increase* the probability of being observed (no "defiers")
  - Or treatment can only *decrease* the probability of being observed (no "compliers")


## Intuition for Lee bounds

- Suppose treatment *increases* observation probability: $M_i(1) \ge M_i(0)$
  - More units observed in the treated group than control group

. . .

- Among observed treated units:
  - Some are "survivors" (would be observed even under control)
  - Some are "compliers" (observed *only because* they got treatment)

. . .

- Among observed control units:
  - **All** are "survivors" (they're observed despite being in control)

. . .

- If we could figure out who the compliers are and remove them - we could get point identification.
  - We **can't** - but we can get the **share** of compliers.
  - And we can **worst-case** which units those happen to be.

## The trimming procedure

- Let $p_1 = Pr(M_i = 1 | D_i = 1)$ and $p_0 = Pr(M_i = 1 | D_i = 0)$
- Under monotonicity with $M_i(1) \ge M_i(0)$: $p_1 \ge p_0$

. . .

- The fraction of treated-and-observed who are "compliers" is:

$$q = \frac{p_1 - p_0}{p_1}$$

. . .

- **Idea:** Trim proportion $q$ of the treated group's outcome distribution
  - Trimming from the **top** gives a **lower bound** on $\tau_{\text{SACE}}$
  - Trimming from the **bottom** gives an **upper bound** on $\tau_{\text{SACE}}$

. . .

- Why does this work?
  - We don't know *which* units are compliers, but we know *how many* there are
  - Trimming the extremes gives us worst-case bounds

## Lee (2009) bounds

- Under **random assignment** and **monotonicity**, the SACE is bounded by:

$$\tau^{LB} \le \tau_{\text{SACE}} \le \tau^{UB}$$

. . .

- **Lower bound**: Trim the top $q$ proportion of treated outcomes

$$\tau^{LB} = \mathbb{E}[Y_i | D_i = 1, M_i = 1, Y_i \le y^{1-q}_1] - \mathbb{E}[Y_i | D_i = 0, M_i = 1]$$

- **Upper bound**: Trim the bottom $q$ proportion of treated outcomes

$$\tau^{UB} = E[Y_i | D_i = 1, M_i = 1, Y_i \ge y^{q}_1] - E[Y_i | D_i = 0, M_i = 1]$$

. . .

- Where $y^{q}_1$ is the $q$-th quantile of the treated outcome distribution


## Properties of Lee bounds

- **Sharp bounds**: Cannot be tightened without additional assumptions
  - Any value within the bounds is consistent with the data and assumptions

. . .

- Bounds **collapse to a point** when:
  - Treatment has no effect on selection ($p_1 = p_0$, so $q = 0$)
  - In this case, SACE = ATE (no selection problem!)
  - But the **monotonicity** assumption is key here!

. . .

- Can construct confidence intervals for bounds using bootstrap or analytical standard errors
  - Note that the **bounds** reflect fundamental uncertainty in mapping from **observed outcomes** to **potential outcomes**
  - Further uncertainty is driven by the usual sources (random sampling/treatment assignment)



