---
title: "Week 13, Part 2: Sensitivity Analysis"
subtitle: "PS 813 - Causal Inference"
author:
  - name: "Anton Strezhnev"
    affiliation: "University of Wisconsin-Madison"
    email: strezhnev@wisc.edu
date: "April 22, 2026"
date-format: "MMMM D, YYYY"
editor:
  render-on-save: true
format:
  revealjs:
    theme: [default, custom.scss]
    slide-number: true
    overview: true
    navigation-mode: vertical
    progress: false
    controls: true
    link-external-newwindow: true
    footer: "PS 813 - University of Wisconsin - Madison"
    html-math-method:
      method: mathjax
      url: https://cdn.jsdelivr.net/npm/mathjax@3/es5/tex-mml-chtml.js
execute:
  cache: true
code:
  execute:
    - r
    - python
embed-resources: true

---

$$
  \require{cancel}
$$

```{r}
#| echo: false
#| warning: false
#| message: false
library(tidyverse)
library(estimatr)
library(broom)
options(digits=3)
```


## Today

::: {.incremental}

- Back to **selection on observables**!
- How can we diagnose our no unobserved confounders assumption?
  - What sorts of confounders would threaten our results (cause them to go to $0$).
- Omitted variable bias formula
  - Allows us to "sign the bias" of a proposed confounder based on the outcome-confounder and outcome-treatment relationships
- Sensitivity analysis
  - How bad of a hypothetical confounder would we need to break the result?
  - Provides a benchmark for any critiques of a selection-on-observables assumption.

:::

## Omitted Variable Bias {.title-slide}

## OVB in linear models

::: {.incremental}

- Suppose we want to identify the effect of $D$ on $Y$ conditional on pre-treatment covariates $\mathbf{X}$. Assume we're willing to assume a linear model for the outcome and that there exists one omitted covariate $Z$

  $$Y = \hat{\tau} D + \mathbf{X}\hat{\beta} + \hat{\gamma} Z + \hat{\epsilon}$$

- What happens if we instead estimate the "restricted" model with $Z$ omitted?

  $$Y = \hat{\tau}_{\text{res}} D + \mathbf{X}\hat{\beta}_{\text{res}} + \hat{\epsilon}_{\text{res}}$$

- What's the relationship between $\hat{\tau}_{\text{res}}$ and $\hat{\tau}$?

:::

## OVB in linear models

::: {.incremental}

- Let's define $D^{\perp \mathbf{X}}$ as the "partialled-out" value of $D$ (the residuals from a regression of $D$ on $X$).
  - Similarly define $Y^{\perp \mathbf{X}}$ as the "partialled-out" value of $Y$ given $X$
- By the Frisch-Waugh-Lovell theorem, we can write any regression coefficient in terms of the "partialled" bivariate regression

  $$\hat{\tau}_{\text{res}} = \frac{\text{cov}(D^{\perp \mathbf{X}}, Y^{\perp \mathbf{X}})}{\text{var}(D^{\perp \mathbf{X}})}$$

- Using our definition of $Y$ (by the linear model)

  $$\hat{\tau}_{\text{res}} = \frac{\text{cov}(D^{\perp \mathbf{X}}, \hat{\tau}D^{\perp \mathbf{X}} + \hat{\gamma}Z^{\perp \mathbf{X}})}{\text{var}(D^{\perp \mathbf{X}})}$$

- Properties of covariance

  $$\hat{\tau}_{\text{res}} = \hat{\tau}\frac{\text{cov}(D^{\perp \mathbf{X}}, D^{\perp \mathbf{X}})}{{\text{var}(D^{\perp \mathbf{X}})}}  + \hat{\gamma}\frac{\text{cov}(D^{\perp \mathbf{X}}, Z^{\perp \mathbf{X}})}{\text{var}(D^{\perp \mathbf{X}})}$$

:::

## OVB in linear models

::: {.incremental}

- Simplifying

  $$\hat{\tau}_{\text{res}} = \hat{\tau}  + \hat{\gamma}\frac{\text{cov}(D^{\perp \mathbf{X}}, Z^{\perp \mathbf{X}})}{\text{var}(D^{\perp \mathbf{X}})}$$

- We can recognize that the last term is the coefficient on $D$ from a regression of $Z$ on $D$ and $X$ (again using FWL) - let's call this $\hat{\delta}$

  $$\hat{\tau}_{\text{res}} = \hat{\tau}  + \hat{\gamma}\hat{\delta}$$

- So the discrepancy between the "restricted"  and "unrestricted" models can be written as the product of two coefficients - the relationship between $Z$ and $Y$ (given $X$) and the relationship between $Z$ and $D$ (given $D$)

  $$\widehat{\text{bias}} = \hat{\gamma}\hat{\delta}$$

:::

## Signing the bias

::: {.incremental}

- Gentzkow, Shapiro, Sinkinson (2011: AER) examine the effect of newspaper entry on political competitiveness in the US counties from 1869 to 1928.
  - Outcome: Presidential/congressional turnout
  - Treatment: Number of new newspapers
  - Finding: More newspapers increase turnout.
- Consider some hypothetical confounders, in what direction would they bias the estimate?
  - **Population growth**: How would population growth affect newspaper entry? How would it affect turnout?
  - **Income growth** How would income growth affect newspaper entry and turnout?
- How would we expect either of these confounders to alter our estimate?

:::

## Sensitivity analyses {.title-slide}

## Sensitivity analysis

::: {.incremental}

- Sensitivity analyses ask the question "how bad of a violation of our identification assumptions would break our result?"
  - We use a parameter (or parameters) to represent the violation and re-do the analysis.
  - Vary the parameter over a (sensible) range - how often do our results appreciably change (e.g. become zero or flip sign)
- **Challenge**: How do we define a suitable sensitivity parameter that has actual interpretability?

:::

## Confounding function

::: {.incremental}

- A general approach to thinking about sensitivity parameters in a binary treatment setting comes from **Blackwell (2014)**
- Define the "confounding function"

  $$q(d, x) = E[Y_i(d) | D_i = d, X_i = x] - E[Y_i(d) | D_i = 1 - d, X_i = x]$$

- The confounding function captures the extent to which the potential outcomes **differ** between a treated unit and a control unit with $X_i = x$.
  - Under ignorability, $q(d, x) = 0$
- We could set the confounding function to have a particular form:
  - For example, $q(d, x) = \alpha$ implies the selection bias is **constant** at all levels of $X_i$

:::

## Sensitivity analyses

::: {.incremental}

- Given a value of $q(d, x)$, we can straightforwardly implement a sensitivity analysis by **de-biasing** the outcome

  $$Y_i^q = Y_i - q(D_i, X_i) \times Pr(1 - D_i | X_i)$$

- Then, run the analysis on $Y_i^q$
  - Vary the sensitivity parameters for $\alpha$ and see what magnitude of confounding is enough to "break" the results.

:::

## Sensitivity analyses

- The intuition for the debiasing comes from our selection-into-treatment bias decomposition. Without covariates:

$$\begin{align*}
E[Y_i(0)] &= E[Y_i(0) |D_i = 0] Pr(D_i = 0) + E[Y_i(0) | D_i = 1] Pr(D_i = 1)\\
&= E[Y_i(0) |D_i = 0] - \bigg(E[Y_i(0) | D_i = 0] - E[Y_i(0) | D_i = 1]\bigg) \times Pr(D_i = 1)\\
&= E[Y_i |D_i = 0] - q(0) \times Pr(D_i = 1)\\
&= E[Y_i^q | D_i = 0]
\end{align*}$$

## Sensitivity analysis in linear models

::: {.incremental}

- An alternative approach is to think about confounding in terms of **two** quantities
  - The relationship between treatment and confounder
  - The relationship between outcome and confounder
- In a linear model setting, we could construct a sensitivity analysis in terms of two parameters:
  - The partial regression coefficient between $Z$ and $Y$, $\hat{\gamma}$
  - The partial regression coefficient between $Z$ and $D$ $\hat{\delta}$
- Slightly annoying since each of these depends on the scale of $Z$ and $Y$ - can we re-write in terms of parameters with the same range irrespective of the outcome?
  - **Cinelli and Hazlett (2020)** provide a reparameterization in terms of the partial $R^2$ of two regressions involving $Z$ (which are always between $0$ and $1$)

:::

## Rewriting the bias

::: {.incremental}

- Start by defining the $R^2_{Z \sim D}$ as the $R^2$ from a regression of $Z$ on $D$.
  - For OLS: $R^2_{Z \sim D} = 1 - \frac{\text{Var}(Z^{\perp D})}{\text{Var}(Z)} = \text{cor}(Z, D)^2 = \left(\frac{\text{cov}(Z, D)}{\text{sd}(Z)\text{sd}(D)}\right)^2$
  - Same thing for the partial $R^2$: $R^2_{Z \sim D | \mathbf{X}} = \text{cor}(Z^{\perp \mathbf{X}}, D^{\perp \mathbf{X}})^2$
- Now write our bias term

  $$\widehat{\text{bias}} = \left(\frac{\text{cov}(D^{\perp \mathbf{X}}, Z^{\perp \mathbf{X}})}{\text{var}(D^{\perp \mathbf{X}})}\right) \left(\frac{\text{cov}(Y^{\perp \mathbf{X}, D}, Z^{\perp \mathbf{X}, D})}{\text{var}(Z^{\perp \mathbf{X}, D})}\right)$$

- Convert covariance to correlation

  $$\widehat{\text{bias}} = \left(\frac{\text{cor}(D^{\perp \mathbf{X}}, Z^{\perp \mathbf{X}}) \text{sd}(Z^{\perp \mathbf{X}})}{\text{sd}(D^{\perp \mathbf{X}})}\right) \left(\frac{\text{cor}(Y^{\perp \mathbf{X}, D}, Z^{\perp \mathbf{X}, D})\text{sd}(Y^{\perp \mathbf{X}, D})}{\text{sd}(Z^{\perp \mathbf{X}, D})}\right)$$

:::

## Rewriting the bias

::: {.incremental}

- Rearrange terms

  $$\widehat{\text{bias}} = \left(\frac{\text{cor}(D^{\perp \mathbf{X}}, Z^{\perp \mathbf{X}}) \text{cor}(Y^{\perp \mathbf{X}, D}, Z^{\perp \mathbf{X}, D})}{\frac{\text{sd}(Z^{\perp \mathbf{X},D})}{\text{sd}(Z^{\perp \mathbf{X}})}}\right) \left(\frac{\text{sd}(Y^{\perp \mathbf{X}, D})}{\text{sd}(D^{\perp \mathbf{X}})}\right)$$

- Square everything

  $$\widehat{\text{bias}}^2 = \left(\frac{\text{cor}(D^{\perp \mathbf{X}}, Z^{\perp \mathbf{X}})^2 \text{cor}(Y^{\perp \mathbf{X}, D}, Z^{\perp \mathbf{X}, D})^2}{\frac{\text{var}(Z^{\perp \mathbf{X},D})}{\text{var}(Z^{\perp \mathbf{X}})}}\right) \left(\frac{\text{var}(Y^{\perp \mathbf{X}, D})}{\text{var}(D^{\perp \mathbf{X}})}\right)$$

- Substitute in the (partial) $R^2$ parameters

  $$\widehat{\text{bias}}^2 = \left(\frac{R^2_{D \sim Z | \mathbf{X}} R^2_{Y \sim Z | \mathbf{X}, D}}{1- R^2_{D \sim Z|\mathbf{X}}}\right) \left(\frac{\text{var}(Y^{\perp \mathbf{X}, D})}{\text{var}(D^{\perp \mathbf{X}})}\right)$$

:::

## Rewriting the bias

::: {.incremental}

- Take the square root to get the absolute bias

  $$|\widehat{\text{bias}}|= \sqrt{\frac{R^2_{D \sim Z | \mathbf{X}} R^2_{Y \sim Z | \mathbf{X}, D}}{1- R^2_{D \sim Z|\mathbf{X}}}}\left(\frac{\text{sd}(Y^{\perp \mathbf{X}, D})}{\text{sd}(D^{\perp \mathbf{X}})}\right)$$

- Some important intuitions
  - The bias is a *product* of the magnitude of the two $R^2$. An unobserved confounder that explains very little of the treatment needs to explain *a lot* of the outcome to induce a sizeable bias (and vice-versa)
  - The bias is smaller when the amount of variation in the outcome given $X$ and $D$ is low (not much $Y$ left to explain)
  - The bias is *amplified* when the variability in $D$ given $X$ is low.

:::

## Illustration: Hazlett (2019)

::: {.incremental}

- **Hazlett (2019; JCR)** examines the impact of exposure to violence on attitudes towards peace in the context of the war in Darfur.
  - Key finding: Refugees with greater exposure to violence are more likely to express support for peace - support for a "war-weariness" theory of attitudes during conflict.
  - Identification strategy: Selection-on-observables conditional on village and gender (plus other covariates).
  - Argues that exposure to violence by pro-government militias across villages was non-random but within-village often indiscriminate.
- How bad does the residual confounding need to be to break the result?

:::

## Illustration: Hazlett (2019)

```{r}
#| echo: true
#| message: false
library(sensemakr)
data('darfur')

darfur.reg <- lm(peacefactor ~ directlyharmed + village + female + age + farmer_dar +
                   herder_dar + pastvoted + hhsize_darfur, data=darfur)
tidy(darfur.reg) %>% filter(term == "directlyharmed")
sd(darfur$peacefactor)
```

## Illustration: Hazlett (2019)

- We would like to generate a plot of how the results would change as we vary the two $R^2$ parameters by calculating the bias across each of the possible parameter values.
  - Need 3 dimensions (each $R^2$ plus the estimate)
  - Can do this manually...but extremely tedious - luckily Cinelli and Hazlett make a great R package `sensemakr`

```{r}
#| echo: true
sensitivity <- sensemakr(model = darfur.reg,
                         treatment = "directlyharmed",
                         benchmark_covariates = "female",
                         kd = 1:3,
                         ky = 1:3,
                         q = 1,
                         alpha = .05,
                         reduce = T)
```

## Illustration: Hazlett (2019)

```{r}
#| echo: false
#| fig-align: center
#| fig-width: 9
#| fig-height: 7
plot(sensitivity)
```

## Illustration: Hazlett (2019)

- "Robustness Value" - consider a confounder that has an equal partial $R^2$ with the outcome and the treatment -- what's the smallest such $R^2$ necessary to drive the result to $0$ (or insignificance).

```{r}
#| echo: true
sensitivity$sensitivity_stats
```

- While the $R^2$ stats do have an intuitive interpretation, there is no *absolute* scale for what constitutes a "robust" vs. "non-robust" result.
  - It depends *also* on how much unexplained variability there is in the outcome.
  - Useful to *benchmark*  the unobserved confounding against other known confounders

## Illustration: Hazlett (2019)

::: {.incremental}

- Consider the previous contour plot - the red points indicate bias under hypothetical confounding that is $1x$, $2x$, and $3x$ as strong as gender
- Why pick gender (versus any other observed confounder)?
  - We have prior theoretical reasons to believe it's strongly associated with both outcome and treatment.
- Caution with *informal* benchmarking - Cinelli and Hazlett (2020) show that just calculating the observed partial $R^2$s for the benchmarks can be inaccurate.
  - Estimates of how $X$ relates to $Y$ may be biased due to omission of $Z$.
  - Also $D$ is a collider.
- C+H (2020) derive formal bounds for the "benchmark" exercise with a set of observed covariates.

:::

## Summary

::: {.incremental}

- Sensitivity analyses are a tool for *arguing*
  - You can always find values of the sensitivity parameters for which the results fail to hold
  - You can always find values of the senstiviity parameters for which the results *do* hold.
- What sensitivity analyses do is describe how *severe* a violation of selection-on-observables needs to be in order to threaten the main conclusions.
  - (e.g.) Smoking and cancer -- even if there were some unobserved confounder it would need to explain an *enormous* amount of variance for us to conclude no effect.
- Norms are developing about sensitivity analyses and reporting in observational designs
  - "Robustness Values" and contour plots
  - How best to "benchmark" -- what to benchmark against
- Don't be surprised if your reviewers start requesting these!

:::
