---
title: "Week 4: Experiments - Covariate Adjustment"
subtitle: "PS 813 - Causal Inference"
author:
  - name: "Anton Strezhnev"
    affiliation: "University of Wisconsin-Madison"
    email: strezhnev@wisc.edu
date: "February 8, 2026"
date-format: "MMMM D, YYYY"
editor:
  render-on-save: true
format:
  revealjs:
    theme: [default, custom.scss]
    slide-number: true
    overview: true
    navigation-mode: vertical
    progress: false
    controls: true
    link-external-newwindow: true
    footer: "PS 813 - University of Wisconsin - Madison"
    html-math-method:
      method: mathjax
      url: https://cdn.jsdelivr.net/npm/mathjax@3/es5/tex-mml-chtml.js
execute:
  cache: true
code:
  execute:
    - r
    - python
embed-resources: true

---

$$
  \require{cancel}
  \DeclareMathOperator*{\argmin}{arg\,min}
$$


$$
 \DeclareMathOperator*{\argmax}{arg\,max}
$$

```{r}
#| echo: false
#| message: false
#| warning: false
library(tidyverse)
library(estimatr)

knitr::opts_chunk$set(digits = 3)
options(digits=3)
```

## Where we've been

::: {.incremental}

- **Causal effects** as contrasts in **potential outcomes**
  - **Average** causal effects as estimands that we can **identify** under assumptions on the experimental design
  
- **Randomized experiments** identify **average treatment effects**
  - Observed treated (control) groups are **representative** of what would happen **on average** if whole sample received treatment (control)
  
- Conditioning on **post-treatment** covariates breaks your experiment
  - Endogenous selection bias - treated group no longer comparable to the controls
  - **For non-compliance** - we'll learn how to identify the **Complier Average Causal Effect**
  
- External validity is a question of **effect heterogeneity**

:::

## Some thoughts on identification assumptions

::: {.incremental}

- (Nearly) every causal identification assumption can be thought of as a variation on these **three ideas**

- **Randomization** - Something is statistically **independent** of something else
  - (e.g.) Ignorability of **treatment** for a conventional randomized experiment
  
- **Effect homogeneity** - We know something about the *individual* treatment effects
  - (e.g.) Monotonicity - effects are only in *one* direction.
  
- **Correct model** - We know the true model for something (e.g. outcome $Y$)
  - Hard to credibly justify in most social science setting, but underpins a lot of "pre-credibility revolution" empirical work even if not explicit.

:::

## This week

::: {.incremental}

- What if we want more **precision** in our experiments
  - We can increase the sample size...but that's expensive!
  - For a **given** sample size, can we do better than complete randomization?
  
- **Covariate adjustment** eliminates variability in $Y$ driven by **non-treatment** characteristics 
  - Adjustment **ex-ante** via stratification/blocking
  - Adjustment **ex-post** via post-stratification
  
- Linear regression in its **agnostic** form.
  - OLS as an estimator of the **Best Linear Predictor**
  - Can we get the model wrong for $E[Y | X]$ and still get consistent estimates of treatment effects? **Yes!**

:::

## Stratification {.title-slide}

## Power and Precision

::: {.incremental}

- In a conventional hypothesis test, we control the **Type I** error - the **false positive rate**
  - **Given** that the null is true, what's the probability that we reject it?
  - We **choose** this number (e.g.) $\alpha = .05$

- But we also want to know the **Type II** error - the **false negative rate**
  - **Given** that the null is **false**, what's the probability that we **fail** to reject it?

- The **false negative rate** in our typical hypothesis test depends on:
  - **The true effect size**
  - **The sampling variance of the estimator**

- For the Neyman variance, what makes it **smaller**?

  $$Var(\hat{\tau}) = \frac{S^2_t}{N_t} + \frac{S^2_c}{N_c}$$

:::

## Covariate adjustment

::: {.incremental}

- How do we reduce the outcome variance?
  - Often we'll **transform** the outcome (e.g. differencing using a pre-treatment measure of that outcome)

- But our most useful technique is to find factors that **predict** the outcome and **eliminate** the variance attributable to those factors.

- Recall from the **law of total variance**

  $$Var(Y) = \underbrace{\mathbb{E}[Var(Y | X)]}_{\text{unexplained by } X} + \underbrace{Var(\mathbb{E}[Y|X])}_{\text{explained by } X}$$
- Adjusting for covariates $X$ in an experiment **eliminates** variance in $Y$ that is attributable to $X$

:::

## Stratification/Blocking

::: {.incremental}

- It is common to build the covariate adjustment directly into the experimental design through **blocking** or **stratification**
- Before assigning treatment, partition the sample that puts the $N$ units into $G$ mutually-exclusive strata
  - With discrete covariates, each bin $g$ is a unique combination of the covariates
  - With continuous covariates, we can coarsen into discrete bins (e.g. age $\to$ "25-40")
  - Each stratum $g$ has $N_g$ units.
- Within **each stratum** treatment is completely randomized
  - $N_{t, g}$ units receive treatment
  - $N_{c, g} = N_g - N_{t, g}$ units receive control.

:::

## Quiz

::: {.incremental}

- If $N_{t, g}/N_g$ is the same across all strata, can I analyze the experiment like I would with complete randomization?

- How about if $N_{t,g}/N_g$ *varies* between strata?

:::

## Estimation

::: {.incremental}

- Our estimator is the **stratified** difference-in-means

- We start by estimating each **stratum-specific** treatment effect $\tau_g$ using the difference-in-means in that stratum

  $$\hat{\tau_g} =  \frac{1}{N_{t,g}}\sum_{i: G_i = g}^N Y_i D_i - \frac{1}{N_{c,g}}\sum_{i: G_i = g}^N Y_i (1 - D_i)$$

- And we can estimate the sampling variance using the within-stratum Neyman variance estimator

  $$\widehat{Var(\hat{\tau_g})} = \frac{s_{t,g}^2}{N_{t,g}} + \frac{s_{c,g}^2}{N_{c,g}}$$

- $s_{t,g}^2$, $s_{c,g}^2$ are the sample variance of $Y$ within the treated group and control group respectively in stratum $g$.

- **Imagine:** We ran $G$ independent mini-experiments and analyzed them separately. Each is unbiased for the conditional ATE (CATE) and has its own standard error.
  - But we care about the **average treatment effect**

:::

## Estimation under block-randomization

::: {.incremental}

- How do we aggregate to get an estimate of the ATE? Take a weighted average by stratum size

  $$\hat{\tau} = \sum_{g = 1}^G \hat{\tau_g} \times \frac{N_g}{N}$$

- And the sampling variance?

  $$\widehat{Var_{\text{strat}}(\hat{\tau})} = \sum_{g = 1}^G \widehat{Var(\hat{\tau_g})}  \times \left(\frac{N_g}{N}\right)^2$$

- When is the variance of the blocked design going to be lower than the variance under complete randomization?
  - When the strata **explain some of the variance** in $Y$ (the population $S^2_{t,g} < S^2_{t}$ and $S^2_{c,g} < S^2_{c}$ )

:::

## Can blocking hurt?

::: {.incremental}

- Long debate over where it's possible to go wrong blocking. **Athey and Imbens (2017)** argue no downside to blocking.
- This answer depends on the framework for inference.
  - **Pashley and Miratrix (2021)** give an extensive review under alternative sampling/inference schemes.

- **Athey and Imbens** result holds under stratified random sampling from the population and equal treatment probability w/in strata.
  - Intuition: In the worst case scenario, stratification is just a two-stage randomization process equivalent to complete randomization.

- This also does not guarantee that the *estimated* standard error will be smaller 
  - With an irrelevant covariate, we will have fewer degrees of freedom as we are estimating multiple parameters. Estimated SEs under stratification
might be higher.

- Athey and Imbens suggest falling back on the conservative **complete randomization** SE,

:::

## Post-stratification

::: {.incremental}

- Suppose we didn't stratify ex-ante but have some covariates that we observe? Can we analyze
*as-though* we had stratified on these?
  - Yes: **Post-stratification**

- Key difference from stratification: Number of treated/control w/in stratum is random and not fixed. Stratum sizes also not fixed.
  - Not as efficient as if we had blocked ex-ante!

- **Miratrix, Sekhon and Yu (2013)**
  - Usually not a problem - relative to blocking ex-ante, the differences in variances are small.
  - Problems with many strata + poorly predictive strata.
  - Unlike the Athey and Imbens setting, benefits not guaranteed, but often doesn't hurt with good covariate choice.

:::

## Example: Nyhan and Reifler (2015)

::: {.incremental}

- **Nyhan and Reifler (2015)** conduct a field experiment on state legislators in November 2012 to study whether politicians reacted to external monitoring of their statements.
  - **Treatment** - Legislators received a letter warning them of the risks of having false statements exposed by fact-checkers
  - **Placebo** - Legislators received a letter warning that their statements would be observed.
  - **Control** - No mailer
  
- Treatment was **block-randomized** 
  - **Exact blocking** on state, political party, legislative chamber and existence of a previous PolitiFact rating
  - Coarsened matching on previous vote share and fundraising
  
- **Outcome**: 
  - Did the legislator receive a negative PolitiFact rating?
  - Was there media coverage of a legislator's inaccurate statements?
  
:::

## Example: Nyhan and Reifler (2015)

- For the main analysis, the paper combines the Placebo/Control conditions.

```{r, echo=T}
factcheck <- haven::read_dta("assets/data/nyhan-reifler-fe.dta")
table(factcheck$treatment)
```

- Let's check balance
  - We know balance is exact on state/party/chamber and whether there is a past fact check.
  - Coarse balance on fundraising and voteshare
  - Other covariates observed but not explicitly "blocked" on (e.g. leadership position)

## Example: Nyhan and Reifler (2015)

```{r bal-love, message = F}
#| echo: false
#| fig-width: 10
#| fig-height: 6
#| fig-align: center
# Create a Love plot to visualize
cobalt::love.plot(
  x = factcheck %>% select(state, gop, chamber, anycheck, partyleader, commleader, fundraising, voteshare),  # Covariate data
  treat = factcheck$treatment,           # Treatment indicator
  weights = factcheck$aw,
  binary = "std",                # Standardize binary variables
  continuous = "std",            # Standardize continuous variables
  s.d.denom = "pooled"           # Use pooled SD for standardization
) + scale_color_discrete(labels = c("Unweighted", "IP-Weighted"))
```

## Example: Nyhan and Reifler (2015)

- The **original paper** estimates the treatment effect on each of the three outcomes using a **weighted** difference-in-means.
  - The weights are negligible but address the slight non-uniformity in treatment assignment across randomization strata.
  
- Here, we replicate the original estimate for the **combined** "any PolitiFact check or LexisNexis news article questioning accuracy" outcome

```{r, echo=T}
dm_est <- tidy(lm_robust(combined ~ treatment, data=factcheck, weights=aw))
dm_est %>% filter(term == "treatment") %>% select(estimate, std.error, statistic, p.value)
```

- But this is kind of a waste of blocking.
  - Why do the blocking if you're not going to incorporate it into the analysis?!
  - Again, the point of blocking in an experiment is not bias reduction but variance reduction!
  - This paper uses a **too conservative** variance estimator.

## Example: Nyhan and Reifler (2015)

- The paper *does* include the randomization strata. Unfortunately **many** of those strata have only a single treated observation.
  - Why is this a problem? We can't estimate the **stratum-specific variance** with a single observation!

- **Advice**: If you're stratifying, make sure to have **at least two** or so observations per condition per stratum
  - Otherwise, if you are pair-randomizing, just take the **within-pair differences** and use a conventional sample mean estimator.
  
- Let's take a look at the strata and see what covariate levels they capture - maybe we can combine some?

```{r, echo=T}
block_summaries <- factcheck %>% group_by(blockdummy) %>% summarize(state = state[1],
                                                                    chamber = chamber[1],
                                                                    gop = mean(gop), 
                                                                    anycheck = mean(anycheck),
                                                                    N_Treated = sum(treatment),
                                                                    N_Control = sum(treatment == 0))
```

## Example: Nyhan and Reifler (2015)

```{r, echo=F, message=F}
DT::datatable(block_summaries, options = list(pageLength = 5, scrollX = TRUE))
```

## Example: Nyhan and Reifler (2015)

- Basically a lot of strata have too few members within a party who are previously fact checked (fact checks are rare). We'll combine some of these strata for the analysis.
  - Could go through and manually merged - to save time, I'll just ignore chamber (it's not that prognostic)
  
- Let's make those strata (`paste()` just concatenates the raw variable values)

```{r, echo=T}
factcheck <- factcheck %>% mutate(groupedblocks = paste(state, gop, anycheck, sep="-"))
factcheck$groupedblocks[factcheck$groupedblocks == "Virginia-1-1"] <- "Virginia-0-1" # Merge some Virginia strata that still have too few units
```

- Now, we can calculate the **stratified** difference-in-means estimate and compute the correct sampling variance!
  - You can do this with `group_by` and `mean()` and `var()` (or for this case, `weighted.mean()` and `weighted.var()`)
  - But it's **so much easier** to do this with a (special) regression.
  - We'll explain why this works later, but for now, let's show you the estimate!
  
## Example: Nyhan and Reifler (2015)
  
- We use the **Lin (2013)** estimator in the `estimatr` package
  - Mathematically equivalent to our stratified difference-in-means estimator from above with a single covariate representing our stratum dummies (we'll show you why later)

```{r, echo=T}
stratified_est <- tidy(estimatr::lm_lin(combined ~ treatment, covariates = ~ groupedblocks, data=factcheck, weights=aw))
stratified_est %>% filter(term == "treatment") %>% select(estimate, std.error, statistic, p.value)
```

- Compare with the case where we use the **conservative** variance estimator that ignores the strata

```{r, echo=T}
dm_est %>% filter(term == "treatment") %>% select(estimate, std.error, statistic, p.value)
```

## Summary

::: {.incremental}

- **Stratification/Blocking** is a tool for reducing the **variance** of your treatment effect estimator
  - **Ex-ante** - Doesn't hurt 
  - **Ex-post** - Usually doesn't hurt if the covariates are predictive enough.
  
- Take care to use an **estimator** that actually leverages the variance-reduction
  - A **stratified** difference-in-means
  - Equivalent to a correctly specified regression! (the **"Lin"** estimator - we'll show you this shortly)
  
- What to block on?
  - Things that **predict** the outcome
  - Surprisingly hard to find good ones! Best ones tend to be prior measures of the outcome.

:::

## Linear Regression {.title-slide}

## Linear Regression

::: {.incremental}

- We often want to adjust for **additional** covariates when trying to estimate a treatment effect
  - In an RCT - **prognostic** covariates that reduce variance in Y
  - In an observational design - both **prognostic** covariates and **confounders** of Y and X
  
- By the law of total expectation, we can write the ATE as an **expectation** over the distribution of CATEs

  $$\tau = \mathbb{E}_X\bigg[\mathbb{E}[Y_i(1) | X_i] - \mathbb{E}[Y_i(0) | X_i]\bigg]$$
  
- Under consistency and **conditional** ignorability (conditional on $X_i$)

  $$E[Y_i(1) | X_i] = E[Y_i | D_i = 1, X_i]$$
  
- Now we have a **statistical** problem, estimating the CEF of $Y_i$ given $X_i$

:::

## Set-up

::: {.incremental}

- We'll denote our vector of observed observations with $Y = \{Y_1, Y_2, Y_3, Y_4, \dotsc\}$
- We observe $K$ predictors for each unit $i$ collected in a matrix $\mathbf{X}$

$$\mathbf{X} = \begin{bmatrix} — & X_1 & — \\ — & X_2 & — \\ & \vdots & \\ — & X_N & — \end{bmatrix}_{N \times K}$$

- We assume an $i.i.d.$ random sample from the target population of interest (though can relax this!)

:::

## Best Linear Predictor (BLP)

::: {.incremental}

- The **Best Linear Predictor** or **population regression** is a function of some input vector $x$

  $$m(x) = x^\prime\beta = \beta_0 + \beta_1 x_1 + \beta_2 x_2 + \beta_3 x_3 + \dotsc$$

- The $\beta$ parameters minimize the **expected prediction error**

  $$\beta = \argmin_b \ \mathbb{E}[(Y_i - X_i^\prime b)^2]$$

- Note that this is a **population** quantity
  - $m(x)$ is an *estimand* like any other estimand we've talked about
  - $m(x)$ and $E[Y|X]$ might differ substantially!

:::

## Best Linear Predictor (BLP)

- Is there a closed form expression for the population regression coefficients $\beta$?
  - Yes - with some algebra (solving for the first order conditions), we get
  
  $$\beta = (\mathbb{E}[X_iX_i^\prime])^{-1}\mathbb{E}[X_iY_i]$$
  
- Recall from intro that in the **bivariate** case, that the expression for the slope coefficient $\beta_1$ can be written as: 

  $$\beta_1 = \frac{Cov(X_i, Y_i)}{Var(X_i)}$$


## Properties

- What have we assumed so far?
  - Finite mean/variance/covariances for $Y$ and $X$
  - Columns of $X$ are linearly independent (no perfect collinearity)
  
## Projection error

- What can we say about the difference between the BLP and $Y$?

  $$e_i = Y_i - m(X_i)$$

- Rewriting, we have

  $$Y_i = X_i^\prime\beta + e_i$$

## Projection error

- We can show that the projection error is mechanically uncorrelated with the predictors

$$\begin{align*}
\mathbb{E}[X_i e_i] &= \mathbb{E}[X_i(Y_i - X_i^\prime\beta)]\\
&= \mathbb{E}[X_iY_i] - \mathbb{E}[X_iX_i^\prime]\beta\\
&= \mathbb{E}[X_iY_i] - \mathbb{E}[X_iX_i^\prime](\mathbb{E}[X_iX_i^\prime])^{-1}\mathbb{E}[X_iY_i]\\
&= \mathbb{E}[X_iY_i] - \mathbb{E}[X_iY_i] = 0\\
\end{align*}$$

- Since $X_{ik} = 1$ is typically one of the regressors (the "intercept"), this also implies $\mathbb{E}[e_i] = 0$

## Projection error

- This means that the covariance of $X_i$ and the projection error $e_i$ is zero!

$$\begin{align*}
Cov(X_i, \epsilon) = \mathbb{E}[X_ie_i] - E[X_i]E[e_i] = 0 - 0 = 0
\end{align*}$$

- Importantly, we have derived these properties on the **Best Linear Predictor** and the **projection error** without making any assumptions on the error itself!
  - This is **mechanically true** by the way we've defined the BLP

## Estimating the BLP

::: {.incremental}

- Remember, $m(X)$ is a *population quantity* - can we come up with an *estimator* in our sample that is consistent for it?
  - **plug-in principle** - Where we have population expectations, plug in their sample equivalents
  
- Our estimand

  $$\beta = (\mathbb{E}[X_iX_i^\prime])^{-1}\mathbb{E}[X_iY_i]$$

- Our estimator

  $$\hat{\beta} = \bigg(\frac{1}{n} \sum_{i=1}^n X_iX_i^\prime\bigg)^{-1}\bigg(\frac{1}{n} \sum_{i=1}^n X_iY_i\bigg)$$

- Often written as:

  $$\hat{\beta} = (\mathbf{X}^\prime\mathbf{X})^{-1}(\mathbf{X}^\prime Y)$$
  
:::

# Estimating the BLP

- Is $\hat{\beta}$ consistent for $\beta$?
- Let's write $\hat{\beta}$ as a function of $\beta$ and the **projection error**

  $$\hat{\beta} = \beta + \bigg(\frac{1}{n} \sum_{i=1}^n X_iX_i^\prime\bigg)^{-1}\bigg(\frac{1}{n} \sum_{i=1}^n X_ie_i\bigg) $$

::: {incremental}

- We can use the **weak law of large numbers**
  
  $$\frac{1}{n} \sum_{i=1}^n X_iX_i^\prime \overset{p}{\to} \mathbb{E}[X_iX_i^\prime]$$
  
  $$\frac{1}{n} \sum_{i=1}^n X_ie_i \overset{p}{\to} \mathbb{E}[X_ie_i] = 0$$

:::

# Estimating the BLP 

::: {.incremental}

- Plugging it all in, we have 

  $$\hat{\beta} \overset{p}{\to} \beta$$

- Ordinary least squares is consistent for the best linear predictor under relatively mild assumptions
  - i.i.d. sample (can weaken this with LLNs for dependent random variables)
  - finite moments (true for most outcomes and regressors you'll work with)
  - no perfect collinearity (you can check if this is a problem!)
  
- What have we *not* assumed
  - Linear CEF
  - Homoskedastic errors
  - Normal outcome
  
:::

# Asymptotic normality

- In large samples, what is the distribution of $\hat{\beta}$?
  
  $$\sqrt{n}(\hat{\beta} - \beta) = \bigg(\frac{1}{n} \sum_{i=1}^n X_iX_i^\prime\bigg)^{-1}\bigg(\frac{1}{\sqrt{n}} \sum_{i=1}^n X_ie_i\bigg)$$

- We know that $\bigg(\frac{1}{n} \sum_{i=1}^n X_iX_i^\prime\bigg)^{-1} \overset{p}{\to} E[X_iX_i^\prime]^{-1}$

# Asymptotic normality

::: {.incremental}

- We can apply the CLT to the other term. We know the expectation is zero, what about the variance?
  
  $$Var(X_ie_i) = \mathbb{E}[X_i e_i (X_i e_i)^\prime] = \mathbb{E}[e_i^2X_i X_i^\prime]$$
  
- By the CLT, we have 

  $$\frac{1}{\sqrt{n}} \sum_{i=1}^n X_ie_i \overset{d}{\to} \mathcal{N}(0, \mathbb{E}[e_i^2X_i X_i^\prime])$$

- Combining with our other convergence result using Slutsky's theorem gives

  $$\sqrt{n}(\hat{\beta} - \beta) \overset{d}{\to} \mathcal{N}\bigg(0, (E[X_iX_i^\prime])^{-1} (\mathbb{E}[e_i^2X_i X_i^\prime]) (E[X_iX_i^\prime])^{-1}\bigg)$$

:::

# Variance estimation

::: {.incremental}

- Again, we have a bunch of population expectations, let's plug in their sample equivalents!

- Our estimand

  $$Var(\hat{\beta}) = \frac{1}{n} (E[X_iX_i^\prime])^{-1} (\mathbb{E}[e_i^2X_i X_i^\prime]) (E[X_iX_i^\prime])^{-1}$$
  
- Our estimator

  $$\widehat{Var(\hat{\beta})} = \frac{1}{n} \bigg(\frac{1}{n} \sum_{i=1}^n X_i X_i^\prime \bigg)^{-1}\bigg(\frac{1}{n} \sum_{i=1}^n \hat{e_i}^2 X_i X_i^\prime \bigg)\bigg(\frac{1}{n} \sum_{i=1}^n X_i X_i^\prime \bigg)^{-1}$$

:::

# Variance estimation

- This is often known as the "sandwich" estimator

$$\widehat{Var(\hat{\beta})} =  (\mathbf{X}^\prime\mathbf{X})^{-1} (X^\prime \hat{\Sigma} \mathbf{X}) (\mathbf{X}^\prime\mathbf{X})^{-1}$$
where $\hat{\Sigma} = \text{diag}(\hat{e_1}^2, \hat{e_2}^2, \dotsc \hat{e_n}^2)$

- It's also referred to as the **heteroskedasticity-consistent** robust variance estimator
  - We have made no assumptions on the variance of $e_i$ -- we're just plugging in the regression residuals.

